### 2010-07-02 doc/guide/options.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

@@

@@

We below group general options of \clasp,
used to configure its global behavior.
%
\begin{description}
\item[\code{--help,-h}]~\\
Print help information and exit.
\item[\code{--version,-v}]~\\
Print version information and exit.
\item[\code{--stats}]~\\
Print (extended) statistic information before termination.
\item[\code{--file,-f \textit{filename}}]~\\
Read from file \code{\textit{filename}} (rather than from standard input).
\item[\code{--number,-n \textit{n}}]~\\
Compute at most~\code{\textit{n}} answer sets,
$\code{\textit{n}}=\code{0}$ standing for compute all answer sets.
\item[\code{--quiet,-q}]~\\
Do not print computed answer sets.
(This is useful for benchmarking.)
\item[\code{--seed \textit{n}}]~\\
Use seed~\code{\textit{n}} for random number generator.
(This may be used to obtain a reproducible behavior
 when running \clasp\ with randomized features.)
\item[\code{--brave}]~\\
Compute the brave consequences (union of all answer sets) of a logic program.
\item[\code{--cautious}]~\\
Compute the cautious consequences (intersection of all answer sets) of a logic program.
\item[\code{--opt-all}]~\\
Compute all optimal answer sets (cf.\ Section~\ref{subsec:gringo:optimize}).
(This is implemented by enumerating~\cite{gekanesc07c} answer sets that are not worse
 than the best one found so far.)
\item[\code{--opt-restart}]~\\
Restart the search from scratch (rather than doing enumeration~\cite{gekanesc07c})
after finding a provisionally optimal answer set.
(This may speed up finding the optimum.)
\item[\code{--opt-value \textit{n1}[,\textit{n2},\textit{n3}...]}]~\\
Initialize objective function(s) to minimize with \code{\textit{n1}[,\textit{n2},\textit{n3}...]}.
Depending on the reasoning mode,
only equitable or strictly better answer sets are computed.
\item[\code{--supp-models}]~\\
Compute supported models~\cite{apblwa88a} (rather than answer sets).
\item[\code{--dimacs}]~\\
Compute models of a propositional formula in DIMACS/CNF format~\cite{dimacs}.
\item[\code{--trans-ext all|choice|weight|no}]~\\
Compile extended rules~\cite{siniso02a} into normal rules of form (\ref{eq:normal:rule}).
Arguments \code{choice} and \code{weight}
state that all ``choice rules'' or all ``weight rules,'' respectively,
are to be compiled into normal rules,
while \code{all} means that both and \code{no} that none of them
are subject to compilation.
\item[\code{--eq \textit{n}}]~\\
Run equivalence reasoning~\cite{gekanesc08a} for \code{\textit{n}} iterations,
$\code{\textit{n}}=\code{-1}$ and $\code{\textit{n}}=\code{0}$ standing for
run to fixpoint or do not run equivalence reasoning, respectively.
\item[\code{--sat-prepro yes|no|\textit{n1}[,\textit{n2},\textit{n3}]}]~\\
Run \emph{SatElite}-like preprocessing~\cite{eenbie05a}
for at most~\code{\textit{n1}} iterations 
($\code{\textit{n1}}=\code{-1}$ standing for run to fixpoint),
using cutoff~\code{\textit{n2}} for variable elimination
($\code{\textit{n2}}=\code{-1}$ standing for no cutoff), and
for no longer than~\code{\textit{n3}} seconds
($\code{\textit{n3}}=\code{-1}$ standing for no time limit).
Arguments \code{yes} and \code{no} mean
$\code{\textit{n1}}=\code{\textit{n2}}=\code{\textit{n3}}=\code{-1}$
(that is, run to fixpoint)
or that \emph{SatElite}-like preprocessing is not to be run at all, respectively.
\item[\code{--rand-watches yes|no}]~\\
Initially choose watched literals randomly (with argument \code{yes})
or systematically (with argument \code{no}).
\end{description}
%
Having introduced the general options of \clasp,
let us note that the options below \code{--dimacs} in the above list
are quite low-level and more or less an issue of fine-tuning.
More important is the fact that virtually all optimization functionalities
are only provided by \clasp\ if the maximum number of answer sets to be
computed is set to~\code{0} (standing for all answer sets),
as it is likely to stop search too early otherwise.
The same applies to computing either brave or cautious consequences
(via one of the flags \code{--brave} and \code{--cautious}).

@@

Finally, let us consider the default command line of \clasp:
%
\begin{lstlisting}[numbers=none]
clasp 1 --trans-ext=no --eq=5 --sat-prepro=no --rand-watches=yes
        --lookback=yes --lookahead=no --heuristic=Berkmin
        --rand-freq=0.0 --rand-prob=no
        --restarts=100,1.5 --shuffle=0,0 --deletion=3.0,1.1,3.0
        --strengthen=all --loops=common --contraction=250
\end{lstlisting}
%
Considering only the most significant defaults,
numeric argument~\code{1} instructs \clasp\ to terminate immediately
after finding an answer set.
Together with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
restart policy.

>>>>

\begin{newstuff}
	\begin{itemize}
		\item stats is not yet implemented
		\item verbose takes level now
		\item debug corresponds to verbose=3
		\item aspils is gone
		\item bindersplit is gone (not worth implementing, can be done in preprocessor)
	\end{itemize}
\end{newstuff}

@@

\begin{newstuff}
	\begin{itemize}
		\item the subsubsections below are probably outdated, benny?
	\end{itemize}
\end{newstuff}

@@

We below group general options of \clasp,
used to configure its global behavior.
%
\begin{description}
\item[\code{--help,-h}]~\\
Print help information and exit.
\item[\code{--version,-v}]~\\
Print version information and exit.
\item[\code{--stats}]~\\
Print (extended) statistic information before termination.
\item[\code{--file,-f \textit{filename}}]~\\
Read from file \code{\textit{filename}} (rather than from the standard input).
\item[\code{--number,-n \textit{n}}]~\\
Compute at most~\code{\textit{n}} answer sets,
$\code{\textit{n}}=\code{0}$ standing for compute all answer sets.
\item[\code{--quiet,-q}]~\\
Do not print computed answer sets.
(This is useful for benchmarking.)
\item[\code{--seed \textit{n}}]~\\
Use seed~\code{\textit{n}} (rather than~\code{1}) for random number generator.
\item[\code{--brave}]~\\
Compute the brave consequences (union of all answer sets) of a logic program.
\item[\code{--cautious}]~\\
Compute the cautious consequences (intersection of all answer sets) of a logic program.
\item[\code{--opt-all}]~\\
Compute all optimal answer sets (cf.\ Section~\ref{subsec:gringo:optimize}).
(This is implemented by enumerating~\cite{gekanesc07c} answer sets that are not worse
 than the best one found so far.)
\item[\code{--opt-restart}]~\\
Restart the search from scratch (rather than doing enumeration~\cite{gekanesc07c})
after finding a provisionally optimal answer set.
(This may speed up finding the optimum.)
\item[\code{--opt-value \textit{n1}[,\textit{n2},\textit{n3}...]}]~\\
Initialize objective function(s) to minimize with \code{\textit{n1}[,\textit{n2},\textit{n3}...]}.
Depending on the reasoning mode,
only equitable or strictly better answer sets are computed.
\item[\code{--supp-models}]~\\
Compute supported models~\cite{apblwa88a} (rather than answer sets).
\item[\code{--dimacs}]~\\
Compute models of a propositional formula in DIMACS/CNF format~\cite{dimacs}.
\item[\code{--trans-ext all|choice|weight|no}]~\\
Compile extended rules~\cite{siniso02a} into normal rules of form (\ref{eq:normal:rule}).
Arguments \code{choice} and \code{weight}
state that all ``choice rules'' or all ``weight rules,'' respectively,
are to be compiled into normal rules,
while \code{all} means that both and \code{no} that none of them
are subject to compilation.
\item[\code{--eq \textit{n}}]~\\
Run equivalence reasoning~\cite{gekanesc08a} for \code{\textit{n}} iterations,
$\code{\textit{n}}=\code{-1}$ and $\code{\textit{n}}=\code{0}$ standing for
run to fixpoint or do not run equivalence reasoning, respectively.
\item[\code{--sat-prepro yes|no|\textit{n1}[,\textit{n2},\textit{n3}]}]~\\
Run \emph{SatElite}-like preprocessing~\cite{eenbie05a}
for at most~\code{\textit{n1}} iterations 
($\code{\textit{n1}}=\code{-1}$ standing for run to fixpoint),
using cutoff~\code{\textit{n2}} for variable elimination
($\code{\textit{n2}}=\code{-1}$ standing for no cutoff), and
for no longer than~\code{\textit{n3}} seconds
($\code{\textit{n3}}=\code{-1}$ standing for no time limit).
Arguments \code{yes} and \code{no} mean
$\code{\textit{n1}}=\code{\textit{n2}}=\code{\textit{n3}}=\code{-1}$
(that is, run to fixpoint)
or that \emph{SatElite}-like preprocessing is not to be run at all, respectively.
\item[\code{--rand-watches yes|no}]~\\
Initially choose watched literals randomly (with argument \code{yes})
or systematically (with argument \code{no}).
\end{description}
%
Having introduced the general options of \clasp,
let us note that the options below \code{--dimacs} in the above list
are quite low-level and more or less an issue of fine-tuning.
More important is the fact that virtually all optimization functionalities
are only provided by \clasp\ if the maximum number of answer sets to be
computed is set to~\code{0} (standing for all answer sets),
as it is likely to stop search too early otherwise.
The same applies to computing either brave or cautious consequences
(via one of the flags \code{--brave} and \code{--cautious}).

@@

Finally, let us consider the default command line of \clasp:
%
\begin{lstlisting}[numbers=none]
clasp 1 --seed=1 --trans-ext=no --eq=5 --sat-prepro=no
        --rand-watches=yes --lookback=yes --lookahead=no
        --heuristic=Berkmin --rand-freq=0.0 --rand-prob=no
        --restarts=100,1.5 --shuffle=0,0 --deletion=3.0,1.1,3.0
        --strengthen=all --loops=common --contraction=250
\end{lstlisting}
%
Considering only the most significant defaults,
numeric argument~\code{1} instructs \clasp\ to terminate immediately
after finding an answer set,
and with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
restart policy.

### 2010-07-02 doc/guide/restrictions.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

In view of function symbols with non-zero arity,
we may be confronted with logic programs over an infinite Herbrand base.
In order to maintain decidability of reasoning tasks,
it is thus important to identify language fragments for which
finite equivalent ground programs are guaranteed to exist.
Level-restricted logic programs constitute such a fragment,
where finiteness is manifested in the requirement that any variable in a rule
must be bound to a finite set of ground terms via a predicate not
subject to positive recursion through that rule.
The notion of level-restrictedness is complemented by stratification,
which by disallowing negative recursion among predicates describes a
class of logic programs having unique answer sets.
The formal definition of level-restricted programs,
which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
is provided in Section~\ref{subsec:lambda}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.

\subsection{Level-Restricted Logic Programs}\label{subsec:lambda}

@@

The fundamental property of a level-restricted
program is that it admits a finite equivalent ground program, not to be confused with
the (full) ground instantiation because $\ground{\mathcal{T}}$ may still be infinite.
Furthermore, level-restrictedness provides a syntactic criterion that is not
difficult to check.
In fact, \gringo\ performs such a check and accepts an input program if it is level-restricted.
Generalizations of level-restrictedness to the rich input language of \gringo\
will be discussed in Section~\ref{sec:language}.

@@

Another relevant restriction is stratification (cf.~\cite{minker88}),
as every stratified (normal) program has a unique answer.
Grounder \gringo\ exploits stratification in two respects:
first, it fully evaluates stratified (sub)programs during grounding, and second,
predicates that are subject to stratification can serve as domain predicates (see below).
As with level-restrictedness, stratification can be characterized in terms
of level mappings, here, witnessing the absence of recursion through ``\code{not}.''
We start by introducing stratification on normal programs, and in a second step,
we consider stratified subprograms.

@@

\begin{example}\label{ex:strat:max}
As the program in Line 1--4 of Example~\ref{ex:strat:prog} is stratified,
the maximal stratified subprogram consists of all rules.
After adding the rule in Line~5,
only the rules in Line~1--3 belong to the maximal stratified subprogram.
The corresponding domain predicates are
$\pred{fruit}/1$,
$\pred{foul}/1$,
$\pred{natural}/1$, and
$\pred{healthy}/1$, 
and the unique answer set for them is
$
Y
=
\{
\code{\pred{fruit}(\const{apple})},\linebreak[1]
\code{\pred{fruit}(\const{peach})},\linebreak[1]
\code{\pred{foul}(\const{peach})},\linebreak[1]
\code{\pred{natural}(\const{apple})},\linebreak[1]
\code{\pred{natural}(\const{peach})},\linebreak[1]
\code{\pred{healthy}(\const{apple})}
\}
$.
\eexample
\end{example}

One can check that the program consisting of Line~1--5
in Example~\ref{ex:strat:prog} is level-restricted, but not stratified.
Conversely, program
\begin{lstlisting}[numbers=none]
nat(0).  nat(s(X)) :- nat(X).
\end{lstlisting}
is stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$),
but not level-restricted
($\lambda(\pred{nat}/1)>\lambda(\pred{nat}/1)$
 needed because of the second rule is impossible).
In fact, the program has a unique but infinite answer set.
Thus,
the program cannot be translated to a finite equivalent ground program,
which is why \gringo\ does not accept it.
Rather, in the first place, a program must be level-restricted in order
to ground it with \gringo.
If level-restrictedness is granted,
the maximal stratified subprogram is exploited in a second stage
to optimize grounding and to make use of domain predicates.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

>>>>

\begin{oldstuff}
In view of function symbols with non-zero arity,
we may be confronted with logic programs over an infinite Herbrand base.
In order to maintain decidability of reasoning tasks,
it is thus important to identify language fragments for which
finite equivalent ground programs are guaranteed to exist.
Level-restricted logic programs constitute such a fragment,
where finiteness is manifested in the requirement that any variable in a rule
must be bound to a finite set of ground terms via a predicate not
subject to positive recursion through that rule.
The notion of level-restrictedness is complemented by stratification,
which by disallowing negative recursion among predicates describes a
class of logic programs having unique answer sets.
The formal definition of level-restricted programs,
which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
is provided in Section~\ref{subsec:lambda}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.
\end{oldstuff}

\begin{newstuff}
In order to apply a grounding algorithm to a logic program,
we have to impose certain restrictions on the input format.
We will allow for a very general class of logic programs,
so called safe programs,
that give us a lot of modeling flexibility.
In fact, \gringo-3.0's new input language is Turing-complete.
This implies that in contrast to previous versions\comment{refer to appendix},
the grounding is no longer guaranteed to be finite
and thus the grounding algorithm will not necessarily halt.
On first sight this might seem to be a disadvantage but is compensated by the fact 
that much more interesting problems, often in a more concise way, can be represented now.
For example, we can run some calculation till a certain fix-point is reached.
This has not been possible before and frequently workarounds were needed, e.g, 
large domains that provide the necessary values to perform the desired calculation.
With safe programs such values can be calculated on demand,
which often prevents space explosion, and also keeps the encoding more compact.
The formal definition of safe programs programs,
which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
is provided in Section~\ref{subsec:safe}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.
\end{newstuff}

\begin{oldstuff}
\subsection{Level-Restricted Logic Programs}\label{subsec:lambda}

@@

The fundamental property of a level-restricted
program is that it admits a finite equivalent ground program, not to be confused with
the (full) ground instantiation because $\ground{\mathcal{T}}$ may still be infinite.
Furthermore, level-restrictedness provides a syntactic criterion that is not
difficult to check.
In fact, \gringo\ performs such a check and accepts an input program if it is level-restricted.
Generalizations of level-restrictedness to the rich input language of \gringo\
will be discussed in Section~\ref{sec:language}.
\end{oldstuff}

\begin{newstuff}
\subsection{Safe Logic Programs}\label{subsec:safe}

The main task of a grounder is to substitute the variables in a logic program~$\Pi$
by terms such that the result is a finite equivalent ground program~$\alt{\Pi}$.%
\footnote{$\Pi$ and~$\alt{\Pi}$ are equivalent if they have the same answer sets.}
Of course, a necessary condition for this is that~$\Pi$ possesses
only finitely many finite answer sets.
Unfortunately, such a property is undecidable in general~\cite{daeigovo01a}.
Instead of semantic properties, grounders do thus impose rather simple syntactic
conditions to guarantee the existence of finite equivalent ground programs
(which are then also computed).
For instance, the \dlv\ system~\cite{dlv03a} requires programs to be \emph{safe} for
establishing finiteness in the absence of functions with non-zero arity and under
limited arithmetic.
Grounder \lparse~\cite{lparseManual} deals with \emph{$\omega$-restricted}
programs~\cite{syrjanen01a}, which are more restrictive than safe programs
but guarantee finiteness also in the presence of functions with non-zero arity.
Previous \gringo~\cite{gescth07a} versions required programs to be \emph{$\lambda$-restricted},
a property more general than $\omega$-restrictedness but likewise applicable
to programs over the same languages.
Finally, recent \gringo\ versions and, e.g., \dlvcomplex~\comment{add reference} use \emph{safe programs}, 
which are a strict super set of \emph{$\lambda$-restricted} programs 
and, in constrast to \dlv\, do not limit arithmetics and also allow for using function symbols.
In this setting, it is no longer possibile to check
whether there is a finite grounding for a given program and 
thus a grounding algorithm need not necessarily hold on some inputs.

A logic program is safe if every variable in each rule occurs in some positive predicate of the rule.
In this case, we can apply \emph{semi-naive evaluation} \comment{add reference} to ground the logic program.
Whenever we ground a rule and see a new head, we can use this head to bind variables
in corresponding positive predicates to obtain new ground rules.
This process is repeated until no more new ground rules will be generated.
The sketched approach is justified by the $\subseteq$-minimality of answer sets
in the sense of Section~\ref{subsec:semantics},
as it allows us to restrict the attention to rules whose bodies are derivable w.r.t.\
some answer set.

\begin{example}\label{ex:safe}
Consider the following logic program:
%
\lstinputlisting{examples/zigzagx.lp}
%
The set~$\mathcal{P}$ of predicates contains
\pred{zig}/$1$, \pred{zag}/$1$,
and \pred{zigzag}/$2$.
The rules~$r$ with head predicates \pred{zig}/$1$ or \pred{zag}/$1$ in Line~1 and~2
are ground ($\vars{r}=\nolinebreak\emptyset$).
Hence, we immediately get the two ground rules
\begin{lstlisting}[firstnumber=1]
zigzag(0).
zigzag(1).
\end{lstlisting}
There are two new heads $\pred{zig}(0)$ and $\pred{zag}(1)$ that can be used to instantiate the third rule
\begin{lstlisting}[firstnumber=3]
zigzag(0,1) :- zig(0), zag(1).
\end{lstlisting}
This allows use to use $\pred{zigzag}(0,1)$ to instantiate the last rule
\begin{lstlisting}[firstnumber=4]
zigzag(1,0) :- zigzag(0,1).
\end{lstlisting}
We get a new head $\pred{zigzag}(1,0)$ that will be used to instantiate the same rule again
\begin{lstlisting}[firstnumber=5]
zigzag(0,1) :- zigzag(1,0).
\end{lstlisting}
At this point, no new heads have been generated and the grounding process stops.
\end{example}

\begin{example}\label{ex:unsafe}
Note that safety is a very intuitive concept.
Consider for example the following unsafe program
\begin{lstlisting}[firstnumber=1]
a(X) :- not b(Y).
\end{lstlisting}
Using an infinite Herbrand base, the above program neither has a finite instantiation nor a finite answer set
and we cannot apply the grounding algorithm sketched above.\comment{Refer to literature that covers such programs?}
\end{example}

The fundamental property of a safe
program is that it admits using the grounding algorithm sketched above 
and that it might yield a finite ground instantiation, not to be confused with
the (full) ground instantiation because $\ground{\Pi}$ may still be infinite.
Furthermore, safety provides a syntactic criterion that is not
difficult to check.
In fact, \gringo\ performs such a check and accepts an input program only if it is safe.
To generalize the notion of safety to the rich input language of \gringo,
we formalize safety using a directed \emph{variable graph} $G_r=(V \cup \{\$\},E)$
w.r.t. a rule~$r$.
The set~$V$ corresponds to the variables that occur in the rule and
the set of edges~$E \subseteq (V \cup \{\$\}) \times V$ connect variables 
or the special start node $\$$ with other variables.
Only positive predicates contribute to edges of the graph.
For every variable~$X$ that occurs in a positive literal of the rule there is an edge~$(\$,X)\in E$.
A rule~$r$ is safe if there is a level mapping $\lambda:V\cup\{\$\}\rightarrow\mathbb{N}$ 
and for every~$X \in V$ exits a~$Y \in V\cup\{\$\}$ 
such that $\lambda(X) < \lambda(Y)$ and $(X,Y) \in E$ for $G_r=(V \cup \{\$\}, E)$.
Finally, a logic program $\Pi$ is safe if all its rules are safe.
In Section~\ref{sec:language}, extensions to this graph will be dicussed
whenever new language constructs are introduced.
\end{newstuff}

@@

Another relevant restriction is stratification (cf.~\cite{minker88}),
as every stratified (normal) program has a unique answer.
Grounder \gringo\ exploits stratification in two respects:
first, it fully evaluates stratified (sub)programs during grounding, and second,
predicates that are subject to stratification can serve as domain predicates (see below).
Stratification can be characterized in terms
of level mappings, here, witnessing the absence of recursion through ``\code{not}.''
We start by introducing stratification on normal programs, and in a second step,
we consider stratified subprograms.

@@

\begin{example}\label{ex:strat:max}
As the program in Line~1--4 of Example~\ref{ex:strat:prog} is stratified,
the maximal stratified subprogram consists of all rules.
After adding the rule in Line~5,
only the rules in Line~1--3 belong to the maximal stratified subprogram.
The corresponding domain predicates are
$\pred{fruit}/1$,
$\pred{foul}/1$,
$\pred{natural}/1$, and
$\pred{healthy}/1$, 
and the unique answer set for them is
$
Y
=
\{
\code{\pred{fruit}(\const{apple})},\linebreak[1]
\code{\pred{fruit}(\const{peach})},\linebreak[1]
\code{\pred{foul}(\const{peach})},\linebreak[1]
\code{\pred{natural}(\const{apple})},\linebreak[1]
\code{\pred{natural}(\const{peach})},\linebreak[1]
\code{\pred{healthy}(\const{apple})}
\}
$.
\eexample
\end{example}

Consider the safe, stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$) program
\begin{lstlisting}[numbers=none]
nat(0).  nat(s(X)) :- nat(X).
\end{lstlisting}
In fact, the program has a unique but infinite answer set.
Thus,
the program cannot be translated to a finite equivalent ground program.
Nevertheless, \gringo\ would accept it, but it will not halt (assuming infinite memory).
Stratified subprograms can be exploited during instantiation
to optimize the grounding, e.g. facts in rule bodies need not be printed, and to make use of domain predicates, e.g., in aggregates~\comment{reference aggregates}.
